Massively Parallel Monte Carlo Methods for Discrete Linear and Nonlinear Systems

### 2012-10-31 doc/prelim/document/chapters/research_proposal/research_proposal.tex
spell check for prelim
adding drekar images

\subsection{Generalization of MCSA for Linear Problems}
\label{subsec:mcsa_generalization}
Previous implementations of MCSA had an entirely physics-based
approach \citep{evans_monte_2009,evans_monte_2012}. In these works,
the transition probabilities and weights were generated directly from
the discretized differential equations instead of a general linear
operator. With this type of approach, transition probabilities and
weight formulations must be rederived for each type of physics
implemented. To reach the broader physics community and facilitate
deployment for more complicated physical systems and multiple physics
problems, MCSA would benefit from an implementation that utilizes the
general operator form of a linear system as in
Eq~(\ref{eq:linear_problem}). Operating on the general form of the
linear system will also greatly simplify the implementation of our
numerical experiments as different equations sets and boundary
conditions can be rapidly implemented within the Panzer library to
generate a complex array of linear problems.

@@

\subsection{FANM Verification}
\label{subsec:fanm_verification}
To verify the FANM method for nonlinear problems, we choose benchmark
solutions for the 2-dimensional, steady, incompressible Navier-Stokes
equations on a rectilinear grid in much the same way as Shadid and
Pawlowski's work on Newton-Krylov methods for the solution of these
equations \citep{shadid_inexact_1997,pawlowski_globalization_2006}. We
define these equations as follows :
\begin{subequations}
  \begin{gather}
    \rho \ve{u} \cdot \nabla \ve{u} - \nabla \cdot \ve{T} - \rho
    \ve{g} = \ve{0}
    \label{eq:ns_momentum}\\
    \nabla \cdot \ve{u} = 0
    \label{eq:ns_continuity}\\
    \rho C_p \ve{u} \cdot \nabla T + \nabla \cdot \ve{q} = 0\:,
    \label{eq:ns_energy}
  \end{gather}
  \label{eq:navier_stokes}
\end{subequations}
where $\rho$ is the fluid density, $\ve{u}$ is the fluid velocity,
$\ve{g}$ gravity, $C_p$ the specific heat capacity at constant
pressure of the fluid and $T$ the temperature of the
fluid. Eq~(\ref{eq:ns_momentum}) provides momentum transport,
Eq~(\ref{eq:ns_continuity}) provides the mass balance, and
Eq~(\ref{eq:ns_energy}) provides energy transport with viscous
dissipation effects neglected. In addition, we close the system with
the following equations:
\begin{subequations}
  \begin{gather}
    \ve{T} = -P \ve{I} + \mu[\nabla \ve{u} + \nabla \ve{u}^T]
    \label{eq:ns_stress_tensor}\\
    \ve{q} = - k \nabla T\:,
    \label{eq:ns_heat_flux}
  \end{gather}
  \label{eq:ns_closure}
\end{subequations}
where $\ve{T}$ is the stress tensor, $P$ is the hydrodynamic pressure,
$\mu$ is the dynamic viscosity of the fluid, $\ve{q}$ is the heat flux
in the fluid, and $k$ is the thermal conductivity of the fluid. This
set of strongly coupled equations possesses both the nonlinearities
and asymmetries that we are seeking for qualification of the FANM
method. Further, physical parameters within these equations can be
tuned to enhance the nonlinearities. We will then apply these
equations to the following three standard benchmark problems.

@@

\subsection{Parallel Performance and Numerical Accuracy Studies for
  the MCSA Method}
\label{subsec:mcsa_studies}
A parallel MCSA implementation and our numerical test framework will
allow us to compare its performance to Krylov methods and allow us to
experiment with the way we combine it with the Neumann-Ulam solver. As
discussed in \S~\ref{subsec:parallel_adjoint}, previous work with
MCSA used a correction generated by the Neumann-Ulam method with a
fairly large uncertainty near 14\%. We hypothesized that if we can
identify the amount of overlap in or Neumann-Ulam experiments required
to eliminate most transitions between adjacent domains owned by
different processors, then perhaps transport to adjacent domains can
be ignored. This will result in a biased estimator for the MCSA
correction that cannot be reproduced, however, these studies will aim
to determine if that bias and lack of reproducibility affects the
numerical results and reproducibility of the MCSA method, yielding a
critical outcome for this work. Furthermore, we will use the
experimental framework to measure the convergence rates for
non-symmetric systems and memory usage of the MCSA algorithm and
compare them to conventional Krylov methods using the same framework.

>>>>

\subsection{Generalization of MCSA for Linear Problems}
\label{subsec:mcsa_generalization}
Previous implementations of MCSA had an entirely physics-based
approach \citep{evans_monte_2009,evans_monte_2012}. In these works,
the transition probabilities and weights were generated directly from
the discretized differential equations instead of a general linear
operator. With this type of approach, transition probabilities and
weight formulations must be re-derived for each type of physics
implemented. To reach the broader physics community and facilitate
deployment for more complicated physical systems and multiple physics
problems, MCSA would benefit from an implementation that utilizes the
general operator form of a linear system as in
Eq~(\ref{eq:linear_problem}). Operating on the general form of the
linear system will also greatly simplify the implementation of our
numerical experiments as different equations sets and boundary
conditions can be rapidly implemented within the Panzer library to
generate a complex array of linear problems.

@@

\subsection{FANM Verification}
\label{subsec:fanm_verification}
To verify the FANM method for nonlinear problems, we choose benchmark
solutions for the 2-dimensional, steady, incompressible Navier-Stokes
equations on a rectilinear grid in much the same way as Shadid and
Pawlowski's work on Newton-Krylov methods for the solution of these
equations \citep{shadid_inexact_1997,pawlowski_globalization_2006}. We
define these equations as follows:
\begin{subequations}
  \begin{gather}
    \rho \ve{u} \cdot \nabla \ve{u} - \nabla \cdot \ve{T} - \rho
    \ve{g} = \ve{0}
    \label{eq:ns_momentum}\\
    \nabla \cdot \ve{u} = 0
    \label{eq:ns_continuity}\\
    \rho C_p \ve{u} \cdot \nabla T + \nabla \cdot \ve{q} = 0\:,
    \label{eq:ns_energy}
  \end{gather}
  \label{eq:navier_stokes}
\end{subequations}
where $\rho$ is the fluid density, $\ve{u}$ is the fluid velocity,
$\ve{g}$ gravity, $C_p$ the specific heat capacity at constant
pressure of the fluid and $T$ the temperature of the
fluid. Eq~(\ref{eq:ns_momentum}) provides momentum transport,
Eq~(\ref{eq:ns_continuity}) provides the mass balance, and
Eq~(\ref{eq:ns_energy}) provides energy transport with viscous
dissipation effects neglected. In addition, we close the system with
the following equations:
\begin{subequations}
  \begin{gather}
    \ve{T} = -P \ve{I} + \mu[\nabla \ve{u} + \nabla \ve{u}^T]
    \label{eq:ns_stress_tensor}\\
    \ve{q} = - k \nabla T\:,
    \label{eq:ns_heat_flux}
  \end{gather}
  \label{eq:ns_closure}
\end{subequations}
where $\ve{T}$ is the stress tensor, $P$ is the hydrodynamic pressure,
$\mu$ is the dynamic viscosity of the fluid, $\ve{q}$ is the heat flux
in the fluid, and $k$ is the thermal conductivity of the fluid. This
set of strongly coupled equations possesses both the nonlinearities
and asymmetries that we are seeking for qualification of the FANM
method. Further, physical parameters within these equations can be
tuned to enhance the nonlinearities. We will then apply these
equations to the following three standard benchmark problems.

@@

\subsection{Parallel Performance and Numerical Accuracy Studies for
  the MCSA Method}
\label{subsec:mcsa_studies}
A parallel MCSA implementation and our numerical test framework will
allow us to compare its performance to Krylov methods and allow us to
experiment with the way we combine it with the Neumann-Ulam solver. As
discussed in \S~\ref{subsec:parallel_adjoint}, previous work with MCSA
used a correction generated by the Neumann-Ulam method with a fairly
large uncertainty near 14\%. We hypothesized that if we can identify
the amount of overlap in our Neumann-Ulam experiments required to
eliminate most transitions between adjacent domains owned by different
processors, then perhaps transport to adjacent domains can be
ignored. This will result in a biased estimator for the MCSA
correction that cannot be reproduced, however, these studies will aim
to determine if that bias and lack of reproducibility affects the
numerical results and reproducibility of the MCSA method, yielding a
critical outcome for this work. Furthermore, we will use the
experimental framework to measure the convergence rates for
non-symmetric systems and memory usage of the MCSA algorithm and
compare them to conventional Krylov methods using the same framework.

### 2012-10-31 doc/prelim/presentation/prelim_presentation.tex
adding drekar images

  \begin{figure}[h!]
    \begin{center}
      \includegraphics[width=4in]{direct_1.png}
    \end{center}
    \caption{\textbf{Direct solution to Poisson Equation.} \textit{1
        history per state, \sn{2.5}{3} total histories. 0.785 seconds
        CPU time.} }
  \end{figure}

@@

  \begin{figure}[h!]
    \begin{center}
      \includegraphics[width=4in]{adjoint_1.png}
    \end{center}
    \caption{\textbf{Adjoint solution to Poisson Equation.}
      \textit{\sn{1}{0} total histories, 0.286 seconds CPU time.} }
  \end{figure}

@@

@@

  \begin{itemize}
  \item No literature observed for parallel Neumann-Ulam solvers
  \item Numerous references for modern parallel Monte Carlo methods in
    reactor physics
  \item MCSA parallelism comes from parallel matrix/vector operations
  \end{itemize}

@@

@@

    \begin{itemize}
    \item Choose a Krylov method to solve for the Newton correction
    \item GMRES with a long recurrence relation obeserved as more
      robust
    \item Generates a monotonically decreasing residual from
      maintaining the optimization and orthogonality conditions
    \end{itemize}

@@

  Domain Overlap Studies for the Parallel Neumann-Ulam Method

  Domain Replication Studies for the Parallel Neumann-Ulam Method

@@

  Parallel Performance and Numerical Accuracy Studies for the MCSA
  Method

  Parallel Performance and Numerical Accuracy Studies for the FANM
  Method

@@

>>>>

  \pause
  \begin{figure}[h!]
    \begin{center}
      \includegraphics[width=4in]{direct_1.png}
    \end{center}
    \caption{\textbf{Direct solution to Poisson Equation.} \textit{1
        history per state, \sn{2.5}{3} total histories. 0.785 seconds
        CPU time.} }
  \end{figure}

@@

  \pause
  \begin{figure}[h!]
    \begin{center}
      \includegraphics[width=4in]{adjoint_1.png}
    \end{center}
    \caption{\textbf{Adjoint solution to Poisson Equation.}
      \textit{\sn{1}{0} total histories, 0.286 seconds CPU time.} }
  \end{figure}

@@

  \begin{itemize}
  \item Work to date has used a physics-based formulation
    \begin{itemize}
    \item Radiation transport equations used as model system
    \item Transition probabilities built from problem-specific
      parameters
    \item Probabilities and weights must be rederived for each new
      equation set
    \end{itemize}
  \item Desire a generalization for all linear operator equations
    \begin{itemize}
    \item Requires a general parallel framework
    \item Requires implementation with a general linear algebra
      framework
    \item Operator, vector, and graph abstractions
    \end{itemize}
  \item Neumann-Ulam solvers and MCSA implemented using the Trilinos
    Petra frameworks
  \item Can be leveraged in modern physics implementations
  \end{itemize}

@@

  \begin{itemize}
  \item No literature observed for parallel Neumann-Ulam solvers
  \item Numerous references for modern parallel Monte Carlo methods in
    reactor physics
  \item Build a strategy for applying modern methods to the
    Neumann-Ulam method
  \item MCSA iteration-level parallelism comes from parallel
    matrix/vector operations
  \end{itemize}

@@

  \pause
  \begin{itemize}
  \item Many multiphysics problems of interest are nonlinear
  \item Segregated methods lack the consitency of fully implicit
    methods
  \item Newton methods often leverage Krylov solvers
    \begin{itemize}
    \item Robust implementations
    \item No operator required
    \item Memory intensive
    \end{itemize}
  \item Monte Carlo methods need the full operator
  \item Automatic construction of the linear operator available
    \begin{itemize}
    \item Ideal for Monte Carlo
    \item Relaxes memory requirements
    \item Potential scaling improvements
    \item Resiliency benefits
    \end{itemize}
  \end{itemize}

@@

    \begin{itemize}
    \item Choose a Krylov method to solve for the Newton correction
    \item GMRES with a long recurrence relation observed as more
      robust
    \item Generates a monotonically decreasing residual from
      maintaining the optimization and orthogonality conditions
    \end{itemize}

@@

  \pause
  Domain Overlap Studies for the Parallel Neumann-Ulam Method
  \begin{itemize}
  \item Correlate domain overlap behavior to linear system properties
  \item Analyze random walk transport with respect to the operator:
    \begin{itemize}
    \item Eigenvalues
    \item Sparsity
    \item Asymmetry
    \end{itemize}
  \item Evaluate communication cost
  \item Evaluate memory cost
  \end{itemize}

  \pause
  Domain Replication Studies for the Parallel Neumann-Ulam Method
  \begin{itemize}
  \item Feasibility from a memory perspective
  \item Impact on scalability
  \item Impact on load balancing
  \end{itemize}

@@

  Parallel Performance and Numerical Accuracy Studies for the MCSA
  Method
  \begin{itemize}
  \item Characterize Neumann-Ulam parameters required for good
    correction
  \item Feasibility of MCSA with full-clip Neumann-Ulam
  \item Performance for assymmetric systems
  \item Memory usage compared to Krylov methods
  \end{itemize}

  \pause
  Parallel Performance and Numerical Accuracy Studies for the FANM
  Method
  \begin{itemize}
  \item Feasibility for problems of interest
  \item Memory usage vs. Newton-Krylov methods
  \item Scalability vs. Newton-Krylov methods
  \end{itemize}

@@

  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.5in]{drekar1.png}
        \end{center}
      \end{figure}
    \end{column}

    \begin{column}{0.5\textwidth}
      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.5in]{drekar2.png}
        \end{center}
      \end{figure}
    \end{column}
  \end{columns}

  \begin{itemize}
  \item Problems of interest are large scale problems in reactor
    physics
  \item CASL has leveraged the Drekar multiphysics code
  \item Coupled fluid flow and heat transfer is required to
    characterize many phenomena
    \begin{itemize}
    \item Departure from nucleate boiling
    \item Grid-to-rod-fretting
    \end{itemize}
  \end{itemize}
